Learning STRIPS action models with classical planning

### 2017-10-16 doc/icaps-18/strips-learning.tex
pages 1-5 updated + new results by Diego

\maketitle
\begin{abstract}
This paper presents a novel approach for learning STRIPS action models from examples that compiles this inductive learning task into a classical planning task. Interestingly the proposed compilation is flexible to different amounts of available input knowledge; the learning examples can range from a set of plans (with their corresponding initial and final states) to just a set of initial and final states where neither actions nor intermediate states are observed. What is more, the compilation can reuse previous knowledge (it is able to start from partially specified action models) and can also be used to validate whether observation of plan executions follow a given STRIPS action model. 
\end{abstract}

\section{Introduction}
Besides plan synthesis~\cite{ghallab2004automated,geffner:book:2013}, planning action models are also useful for plan/goal recognition~\cite{ramirez2012plan}. In both tasks, off-the-shelf planners require reasoning about action models that correctly and completely capture the possible world transitions. Unfortunately, building such planning action models is complex, even for planning experts, and this knowledge acquisition bottleneck limits the potential of automated planning~\cite{kambhampati:modellite:AAAI2007}.

On the other hand, Machine Learning (ML) techniques have shown able to compute a wide range of different kinds of models from examples~\cite{michalski2013machine}. The application of inductive ML techniques for learning planning STRIPS action models from examples is not straightforward though:
\begin{itemize}
\item The inputs to ML algorithms usually are finite vectors that encode the value of fixed features for a given set of objects. The input for learning planning action models traditionally are sets of observations of plan executions (each with possibly a different length).
\item The traditional output of off-the-shelf ML techniques is a scalar value (an integer, in the case of classification tasks, or a real value, in the case of regression tasks). When learning STRIPS action models, the output is not a scalar but the sets of preconditions and effects of each action that define the possible state transitions of a given planning domain. 
\end{itemize}

Learning STRIPS action models is a well-studied problem with sophisticated algorithms, like {\sc ARMS}~\cite{yang2007learning}, {\sc SLAF}~\cite{amir:alearning:JAIR08} or {\sc LOCM}~\cite{cresswell2013acquiring} that do not require full knowledge of the intermediate states traversed by the example plans. Motivated by recent advances on learning different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016hierarchical,segovia2017generating}, this paper introduces an innovative approach for learning STRIPS action models that can be defined as a classical planning compilation (so an off-the-shelf planner can be used to address the inductive learning task). In addition, the compilation approach presents the following benefits:
\begin{enumerate}
\item It is flexible to different amounts of available input knowledge. The learning examples can range from a set of plans (with their corresponding initial and final states) to just a set of initial and final states where no actions or intermediate states are observed.
\item Can exploit previous knowledge about the action model in the form of partially specified action models.
\item Can be used to validate whether observed plan executions are valid for a given STRIPS action model.
\end{enumerate}

The paper is organized as follows. The first section presents the classical planning model, its extension to conditional effects (a requirement for the proposed compilation) and the STRIPS action model. The second section formalizes the task of learning STRIPS action models from observations of plan executions with regard to different amounts of input knowledge. The third section describes our approach for addressing these inductive learning tasks by compiling them into classical planning. The fourth section reports the collected data during the evaluation of the performance of our approach. Finally, the last section discusses the strengths and weaknesses of our approach and proposes some future work.

\section{Background}
This section defines the planning models used on this work.

@@

A {\em classical planning frame} is a tuple $\Phi=\tup{F,A}$, where $F$ is a set of fluents and $A$ is a set of actions. Each action $a\in A$ has a set of literals $\pre(a)\in\mathcal{L}(F)$, called {\em preconditions}, a set of negative effects $\eff^-(a)\in\mathcal{L}(F)$, and a set of positive effects $\eff^+(a)\in\mathcal{L}(F)$. An action $a\in A$ is applicable in state $s$ iff $\pre(a)\subseteq s$, and the result of applying $a$ in $s$ is a new state $\theta(s,a)=(s\setminus \neg\eff^-(a))\cup\eff^+(a)$.

@@

\subsection{Classical planning with conditional effects}
Our approach for leaning STRIPS action models is compiling this leaning task into a classical planning task with conditional effects. We use conditional effects because they allow us to compactly define actions whose particular effects depend on the current state. Many classical planners cope with conditional effects without compiling them away. In fact, the support of conditional effects was a requirement for participating at the IPC-2014~\cite{vallati:IPC:AIM2015}.

@@

An action $a\in A$ is applicable in state $s$ if and only if $\pre(a)\subseteq s$, and the resulting set of {\em triggered effects} is
\[
\eff(s,a)=\bigcup_{C\rhd E\in\cond(a),C\subseteq s} E,
\]
i.e.~effects whose conditions hold in $s$. The result of applying $a$ in $s$ is a new state $\theta(s,a)=(s\setminus \neg \eff^-(s,a))\cup\eff^+(s,a)$, where $\eff^-(s,a)$ and $\eff^+(s,a)$ are the negative and positive effects in $\eff(s,a)$.

\subsection{The STRIPS action model}
In this work we address the task of learning  action  schema  in  the  STRIPS language~\cite{fikes1971strips}. With this regard, we assume that the fluents in $F$ are instantiated from predicates, as in PDDL~\cite{mcdermott1998pddl,fox2003pddl2}. There exists a set of predicates $\Psi$, each $p\in\Psi$ with an argument list of arity $ar(p)$. Given a set of objects $\Omega$, the set of fluents $F$ is then induced by assigning objects in $\Omega$ to the arguments of predicates in $\Psi$, i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$ s.t. $\Omega^k$ is the $k$-th Cartesian power of $\Omega$.

@@

Let $\Omega_v$ be a new set of objects, $\Omega\cap\Omega_v=\emptyset$, that represents {\em variable names}. The cardinality of this set, $|\Omega_v|$, is given by the maximum arity of an action in the given planning frame. For instance, in a three-block blocksworld $\Omega=\{block_1, block_2, block_3\}$ while $\Omega_v=\{v_1, v_2\}$ because the operators {\small\tt stack} and {\small\tt unstack} are the ones with the maximum arity (two parameters each).

Let us define a new set of fluents $F_{v}$ that results instantiating $\Psi$ but using only the {\em variable objects} in $\Omega_v$. In the blocksworld this set is defined as $F_v$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$),on($v_1,v_2$), on($v_2,v_1$),on($v_2,v_2$)\}}.

We are now ready to define a STRIPS action model $\Xi$ as a set of operator schema $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$:
\begin{itemize}
\item $head(\xi)=\tup{name(\xi),pars(\xi)}$, represents the operator {\em header} defined by its corresponding action name and a enumeration of the variable names given by the arity of the operator, $pars(\xi)\in\Omega_v^{ar(\xi)}$. The headers for a 4-operator blocksworld are then {\small\tt pickup($v_1$), putdown($v_1$), stack($v_1,v_2$)} and {\small\tt unstack($v_1,v_2$)}.
\item The preconditions $pre(\xi)\subseteq F_v$, the negative effects $del(\xi)\subseteq F_v$, and the positive effects $add(\xi)\subseteq F_v$ such that, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$.
\end{itemize}

\section{Learning STRIPS action models}
Learning STRIPS action models from fully available input knowledge, i.e. a set of plans where the {\em pre-} and {\em post-states} of every action in a plan are available, is straightforward. In this case, the operators schema are derived lifting the literals that change between the pre and post-state of the corresponding action executions. Likewise, preconditions are derived lifting the minimal set of literals that appears in all the pre-states that correspond to the same operator.

@@

\subsection{Learning from labeled plans}
This learning task is formalized as $\Lambda=\tup{\Psi,\Pi,\Sigma}$: 
\begin{itemize}
\item $\Psi$, the set of predicates that define the abstract state space of a given planning domain. 
\item $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$, the given set of example plans s.t. each plan $\pi_t=\tup{a_1^t, \ldots, a_n^t}$, {\small $1\leq t\leq \tau$}, is an action sequence that induces a state sequence $\tup{s_0^t, s_1^t, \ldots, s_n^t}$ such that, for each {\small $1\leq i\leq n$}, $a_i^t$ is applicable in the state $s_{i-1}^t$ and generates the successor state $s_i^t=\theta(s_{i-1}^t,a_i^t)$.
\item $\Sigma=\{\sigma_1,,\ldots,\sigma_{\tau}\}$, is a set of labels s.t. each plan $\pi_t$, {\small $1\leq t\leq \tau$}, has an associated label $\sigma_t=(s_0^t,s_{n}^t)$ where $s_{n}^t$ is the state resulting from executing the plan $\pi_t$ starting from the state $s_0^t$. 
\end{itemize}

To illustrate this, Figure~\ref{fig:lexample}, shows the contain of $\Psi$, $\Pi=\{\pi_1\}$ and $\Sigma=\{\sigma_1\}$ for a $\Lambda$ task that aims learning the STRIPS action model of the blocksworld. The shown learning task has a single learning example that corresponds to observing a plan execution for inverting a tower of four blocks.

@@

 \caption{\small Learning task in the blocksworld with a single learning example that corresponds to inverting a tower of four blocks.}
\label{fig:lexample}
\end{figure}

A solution to the learning task $\Lambda$ is a set of STRIPS operator schema $\Xi$ (with one schema $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$ for each action with different name in the example plans) that is compliant with the predicates in $\Psi$, the example plans $\Pi$, and their labels $\Sigma$. With this regard, the operator schema shown in Figure~\ref{fig:stack} is compliant with the learning task illustrated by the Figure~\ref{fig:lexample}.

\subsection{Learning from initial/final states}
Here we reduce the amount of input knowledge provided to the learning task. Now $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is replaced by $\Pi'=\{|\pi_1|,\ldots,|\pi_{\tau}|\}$ i.e. $\Pi'$ that does not contain a set of plans but the number of actions of each plan. The learning task is hence redefined as $\Lambda'=\tup{\Psi,\Pi',\Sigma}$. While the previous learning task, $\Lambda$, corresponds to watching an agent acting in the world, this new learning task $\Lambda'$ can be understood as watching only the result of its plan execution but knowing the number of actions performed by the agent. In the case of the learning example shown in Figure~\ref{fig:lexample}, $|\pi_1|=8$.

Finally, we can go a step further and redefine a third learning task $\Lambda''=\tup{\Psi,\Sigma}$ that corresponds to watching only the results of the plan executions. In this case no information about the executed plans is given. A solution to the $\Lambda''$ learning task is a set of operator schema $\Xi$ that is compliant only with the predicates in $\Psi$, and the given set of initial and final states $\Sigma$.

In these two cases $\Lambda'$ and $\Lambda''$, a solution must not only synthesize a possible action models but also the actions could have produced the given labels (this information about the actions is no longer given in the learning examples). However, we assume that at the headers of the operators schema are known.

\section{Learning STRIPS action models with planning}
Our approach for addressing a learning task $\Lambda$, $\Lambda'$ or $\Lambda''$, is compiling it into a classical planning task with conditional effects $P_{\Lambda}$, $P_{\Lambda'}$ or $P_{\Lambda''}$. The intuition behind these compilations is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item Programs the STRIPS action model $\Xi$. For each $\xi\in\Xi$, the solution plan determines the literals that belong to the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item Validates the programmed the STRIPS action model $\Xi$ using the given input knowledge ($\Pi$ if available, and the given set of labels $\Sigma$).  At every learning example {\small $1\leq t\leq \tau$}, then $\Xi$ is used to produce a final state $s_{n}^t$ starting from the corresponding initial state $s_0^t$. If information about the corresponding plan $\pi_t$ is available, then it is used to constrain the validation of the programmed STRIPS action model $\Xi$.
\end{enumerate}

Figure~\ref{fig:plan} shows a classical plan for solving a learning task of the $\Lambda$ class. The shown classical plan programs and validates the operator schema {\tt\small stack} from the blocksworld using the plan $\pi_1$ and label $\sigma_1$ shown in Figure~\ref{fig:lexample}. In particular the steps $[0,8]$ are the actions for programming the preconditions of the operator, steps $[9,13]$ are the actions for programming the operator effects and steps $[14,22]$ are the actions for validating the programmed operator. The details of the compilation are given in the remaining of the section.

\begin{figure}[hbt]
\begin{scriptsize}
\begin{verbatim}
     0 : (program_pre_clear_stack_v1)
     1 : (program_pre_handempty_stack)
     2 : (program_pre_holding_stack_v2)
     3 : (program_pre_on_stack_v1_v1)
     4 : (program_pre_on_stack_v1_v2)
     5 : (program_pre_on_stack_v2_v1)
     6 : (program_pre_on_stack_v2_v2)
     7 : (program_pre_ontable_stack_v1)
     8 : (program_pre_ontable_stack_v2)
     9 : (program_eff_clear_stack_v1)
    10 : (program_eff_clear_stack_v2)
    11 : (program_eff_handempty_stack)
    12 : (program_eff_holding_stack_v1)
    13 : (program_eff_on_stack_v1_v2)
    14 : (apply_unstack a b i1 i2)
    15 : (apply_putdown a i2 i3)
    16 : (apply_unstack b c i3 i4)
    17 : (apply_stack b a i4 i5)
    18 : (apply_unstack c d i5 i6)
    19 : (apply_stack c b i6 i7)
    20 : (apply_pickup d i7 i8)
    21 : (apply_stack d c i8 i9)
    22 : (validate_1)
\end{verbatim}
\end{scriptsize}
 \caption{\small Example of a plan for programing and validating the operator schema unstack using the plan $\pi_1$ and label $\sigma_1$ shown in Figure~\ref{fig:lexample}.}
\label{fig:plan}
\end{figure}

@@

Given a learning task $\Lambda''=\tup{\Psi,\Sigma}$ the compilation outputs a classical planning task $P_{\Lambda''}=\tup{F_{\Lambda''},A_{\Lambda''},I_{\Lambda''},G_{\Lambda''}}$ such that:
\begin{itemize}
\item $F_{\Lambda''}$ extends $F$ with:
\begin{itemize}
\item Fluents representing the programmed action model $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v$ and $\xi \in \Xi$). If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative effect/positive effect in the STRIPS operator schema $\xi\in \Xi$. 
\item A fluent $mode_{prog}$ indicating whether the operator schema are being programmed or they are already programmed operators, so are being validated in the learning examples. 
\item Fluents $\{test_t\}_{1\leq t\leq \tau}$, indicating the learning example where the programmed action model is being validated.
\end{itemize}
\item $I_{\Lambda''}$, contains the fluents from $F$ that encode $s_0^1$ (i.e. the initial state of the first learning example), every fluent $pre_f(\xi)\in F_{\Lambda}$ (note that our compilation assumes that initially, any operator schema, contains all the possible preconditions, no negative effect and no positive effect) and the $mode_{prog}$ set to true. 
\item $G_{\Lambda''}=\{test_t\}$, {\small $1\leq t\leq \tau$}, indicates that the programmed action model is validated in all the learning examples.
\item $A_{\Lambda''}$ contains actions of three different kinds:
\begin{enumerate}
\item The actions for programming the operator schema. This set includes:
\begin{itemize}
\item The actions for removing a {\em precondition} $f\in F_v$ from the action schema $\xi\in\Xi$.

@@

The binding of the operator schema is done implicitly by order of appearance, i.e. the variables in $pars(\xi)$ are bound to the objects in $\omega$ appearing in the same position. To illustrate this, Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying an already programmed operator schema $stack$.

@@

\begin{proof}[Proof sketch]
Once the preconditions of an operator schema $\Xi$ are programmed, they cannot be altered. The same happens with the positive and negative effects that define an operator schema $\xi \in \Xi$ (besides they can only be programmed after the operator preconditions are programmed, because of the preconditions in the $programPre_{f,\xi}$ actions). Furthermore, once the operator schema are programmed they can only be applied because of the value of the $mode_{prog}$ fluent. To solve the classical planning task $P_{\Lambda''}$, the only way of achieving a goal $\{test_t\}$,{\small $1\leq t\leq \tau$} is by executing an applicable sequence of programmed operator schema that achieves the final state $s_n^t$ defined by the associated label $\sigma_t$, starting from the initial $s_0^t$ defined in this same label. If this is done for all the input examples of the learning task (for all the labels $\sigma_t$, {\small $1\leq t\leq \tau$}), it means that the programmed action model $\Xi$ is compliant with all the input knowledge and hence, it is a solution to the action model learning task.
\end{proof}

>>>>

\maketitle
\begin{abstract}
This paper presents a novel approach for learning STRIPS action models from examples that compiles this inductive learning task into a classical planning task. Interestingly, the compilation approach is flexible to different amounts of available input knowledge; the learning examples can range from a set of plans (with their corresponding initial and final states) to just a set of initial and final states where neither actions nor intermediate states are given. What is more, partially specified action models can be used to feed the compilation and the compilation can also be used to validate whether certain observations of plan executions follow a given STRIPS action model.
\end{abstract}

\section{Introduction}
Besides {\em plan synthesis}~\cite{ghallab2004automated,geffner:book:2013}, planning action models are also useful for {\em plan/goal recognition}~\cite{ramirez2012plan}. At both planning tasks, an automated planner is required to reason about action models that correctly and completely capture the possible world transitions. Unfortunately, building planning action models is complex, even for planning experts, and this knowledge acquisition task is a bottleneck that limits the potential of automated planning~\cite{kambhampati:modellite:AAAI2007}.

On the other hand, Machine Learning (ML) has shown to be able to compute a wide range of different kinds of models from examples~\cite{michalski2013machine}. The application of inductive ML to the learning of STRIPS action models, the vanilla action model for planning~\cite{fikes1971strips}, is not straightforward though:
\begin{itemize}
\item The inputs to ML algorithms (the learning data) usually are finite vectors that encode the value of fixed features for a given set of objects. The input for learning planning action models traditionally are sets of observations of plan executions (each with possibly a different length).
\item The traditional output of off-the-shelf ML techniques is a scalar value (an integer, in the case of classification tasks, or a real value, in the case of regression tasks). When learning STRIPS action models the output are, for each action, the sets of preconditions, negative and positive effects, that define the possible state transitions of a given planning domain. 
\end{itemize}

The learning of STRIPS action models is a well-studied problem with sophisticated algorithms, like {\sc ARMS}~\cite{yang2007learning}, {\sc SLAF}~\cite{amir:alearning:JAIR08} or {\sc LOCM}~\cite{cresswell2013acquiring} that do not require full knowledge of the intermediate states traversed by the example plans. Motivated by recent advances on learning different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016hierarchical,segovia2017generating}, this paper introduces an innovative approach for learning STRIPS action models that can be defined as a classical planning compilation. A compilation approach is appealing because an off-the-shelf planner can be used to address the inductive learning task opening the door to the bootstrapping of action models in planning. In addition, a compilation approach presents the following contributions:
\begin{enumerate}
\item It is flexible to different amounts of available input knowledge. The learning examples can range from a set of plans (with their corresponding initial and final states) to just a set of initial and final states where no actions or intermediate states are observed.
\item Can exploit previous knowledge about the action model in the form of partially specified action models. In the extreme, it can be used to validate whether observed plan executions are valid for a given STRIPS action model.
\end{enumerate}

The paper is organized as follows. The first section presents the classical planning model, its extension to conditional effects (which is a requirement for the proposed compilation) and formalizes the STRIPS action model, the output of the addressed learning task. The second section formalizes the task of learning STRIPS action models with regard to different amounts of input knowledge (observations of plan executions). The third section describes our approach for addressing the learning of STRIPS action models by compiling it into classical planning and how previous knowledge can be introduced in the compilation. The fourth section reports the collected data during the empirical evaluation of the compilation approach. Finally, the last section discusses the strengths and weaknesses of the compilation approach for learning action models and proposes some opportunities of future research.

\section{Background}
This section defines the planning models used on this work and the STRIPS action model, output of the addressed learning task.

@@

A {\em classical planning frame} is a tuple $\Phi=\tup{F,A}$, where $F$ is a set of fluents and $A$ is a set of actions. Each action $a\in A$ has a set of literals $\pre(a)\in\mathcal{L}(F)$, called {\em preconditions}, a set of effects $\eff(a)\in\mathcal{L}(F)$. An action $a\in A$ is applicable in state $s$ iff $\pre(a)\subseteq s$, and the result of applying $a$ in $s$ is a new state $\theta(s,a)=(s\setminus \neg\eff(a))\cup\eff(a)$.

@@

\subsection{Classical planning with conditional effects}
Our approach for leaning STRIPS action models is compiling this leaning task into a classical planning task with conditional effects. Conditional effects allow us to compactly define actions whose particular effects depend on the current state. Many classical planners cope with conditional effects without compiling them away. In fact, the support of conditional effects was a requirement for participating at the IPC-2014~\cite{vallati:IPC:AIM2015}.

@@

An action $a\in A$ is applicable in state $s$ if and only if $\pre(a)\subseteq s$, and the resulting set of {\em triggered effects} is
\[
\eff(s,a)=\bigcup_{C\rhd E\in\cond(a),C\subseteq s} E,
\]
i.e.~effects whose conditions hold in $s$. The result of applying $a$ in $s$ is a new state $\theta(s,a)=(s\setminus \neg \eff(s,a))\cup\eff(s,a)$.

\subsection{The STRIPS action model}
This work addresses the learning  action schema in  the  STRIPS language~\cite{fikes1971strips}. To define the output of this learning task, we assume that the fluents in $F$ are instantiated from predicates, as in PDDL~\cite{mcdermott1998pddl,fox2003pddl2}. There exists a set of predicates $\Psi$, each $p\in\Psi$ with an argument list of arity $ar(p)$. Given a set of objects $\Omega$, the set of fluents $F$ is then induced by assigning objects in $\Omega$ to the arguments of predicates in $\Psi$, i.e.~$F=\{p(\omega):p\in\Psi,\omega\in\Omega^{ar(p)}\}$ s.t. $\Omega^k$ is the $k$-th Cartesian power of $\Omega$.

@@

Let $\Omega_v$ be a new set of objects, $\Omega\cap\Omega_v=\emptyset$, that represents {\em variable names}. The cardinality of this set, $|\Omega_v|$, is given by the maximum arity of an action in a given planning frame $\Phi$. For instance, in a three-block blocksworld $\Omega=\{block_1, block_2, block_3\}$ while $\Omega_v=\{v_1, v_2\}$ because the operators {\small\tt stack} and {\small\tt unstack} are the ones with the maximum arity (two parameters each). Let us define also a new set of fluents $F_{v}$ that results instantiating $\Psi$ but using only the {\em variable objects} in $\Omega_v$. In the blocksworld $F_v$={\small\tt\{handempty, holding($v_1$), holding($v_2$), clear($v_1$), clear($v_2$), ontable($v_1$), ontable($v_2$), on($v_1,v_1$),on($v_1,v_2$), on($v_2,v_1$),on($v_2,v_2$)\}}.

We are now ready to define a STRIPS action model $\Xi$ as a set of operator schema $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$ such that:
\begin{itemize}
\item $head(\xi)=\tup{name(\xi),pars(\xi)}$, is the operator {\em header} defined by its corresponding action name and a enumeration of the variable names bound by the arity of the operator, $pars(\xi)\in\Omega_v^{ar(\xi)}$. The headers for a 4-operator blocksworld are {\small\tt pickup($v_1$), putdown($v_1$), stack($v_1,v_2$)} and {\small\tt unstack($v_1,v_2$)}.
\item The preconditions $pre(\xi)\subseteq F_v$, the negative effects $del(\xi)\subseteq F_v$, and the positive effects $add(\xi)\subseteq F_v$ such that, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$.
\end{itemize}

\section{Learning STRIPS action models}
Learning STRIPS action models from fully available input knowledge, i.e. a set of plans where the {\em pre-} and {\em post-states} of every action in a plan are available, is straightforward. In this case, a STRIPS operator schema is derived by lifting the literals that change between the pre and post-state of the corresponding action executions. Likewise, the preconditions are derived lifting the minimal set of literals that appears in all the pre-states that correspond to the same operator schema.

@@

\subsection{Learning from labeled plans}
This learning task is formalized as $\Lambda=\tup{\Psi,\Pi,\Sigma}$: 
\begin{itemize}
\item $\Psi$, the set of predicates that define the abstract state space of a given planning domain. 
\item $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$, the given set of example plans s.t. each plan $\pi_t=\tup{a_1^t, \ldots, a_n^t}$, {\small $1\leq t\leq \tau$}, is an action sequence that induces the corresponding state sequence $\tup{s_0^t, s_1^t, \ldots, s_n^t}$ such that, for each {\small $1\leq i\leq n$}, $a_i^t$ is applicable in the state $s_{i-1}^t$ and generates the successor state $s_i^t=\theta(s_{i-1}^t,a_i^t)$.
\item $\Sigma=\{\sigma_1,,\ldots,\sigma_{\tau}\}$, is a set of labels s.t. each plan $\pi_t$, {\small $1\leq t\leq \tau$}, has an associated label $\sigma_t=(s_0^t,s_{n}^t)$ where $s_{n}^t$ is the state resulting from executing the plan $\pi_t$ starting from the state $s_0^t$. 
\end{itemize}

To illustrate this, Figure~\ref{fig:lexample} shows an example of a task for learning the STRIPS action model of the blocksworld. Figure~\ref{fig:lexample} shows the content of $\Psi$, $\Pi=\{\pi_1\}$ and $\Sigma=\{\sigma_1\}$, this learning task has a single learning example, that corresponds to observing a plan execution for inverting a tower of four blocks.

@@

 \caption{\small Example of a task for learning the STRIPS action model of the blocksworld from a single learning example.}
\label{fig:lexample}
\end{figure}

A solution to a learning task $\Lambda$ is a set of STRIPS operator schema $\Xi$ (with one schema $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$ for each action with a different name in the example plans) that is compliant with the predicates in $\Psi$, the example plans $\Pi$, and their labels $\Sigma$.

\subsection{Learning from initial/final states}
Here we reduce the amount of input knowledge provided to the learning task. First, $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is replaced by $\Pi'=\{|\pi_1|,\ldots,|\pi_{\tau}|\}$ i.e. $\Pi'$ that does not contain a set of plans but the number of actions of each plan. The learning task is hence redefined as $\Lambda'=\tup{\Psi,\Pi',\Sigma}$. While the previous learning task, $\Lambda$, corresponds to watching an agent acting in the world, this new learning task $\Lambda'$ can be understood as watching only the result of its plan execution but knowing the number of actions performed by the agent. In the case of the learning example shown in Figure~\ref{fig:lexample}, $|\pi_1|=8$.

Finally, we go one step further and redefine a third learning task $\Lambda''=\tup{\Psi,\Sigma}$ that corresponds to watching only the results of the plan executions. In this case no information about the executed plans is given. A solution to the $\Lambda''$ learning task is a set of operator schema $\Xi$ that is compliant only with the predicates in $\Psi$, and the given set of initial and final states $\Sigma$.

In these two cases $\Lambda'$ and $\Lambda''$, a solution must not only determine a possible STRIPS action model but also the plans that explain the given labels using the learned model (this information about the actions is no longer given in the learning examples). However, we assume that the headers of the operators schema are known.

\section{Learning STRIPS action models with planning}
Our approach for addressing a learning task $\Lambda$, $\Lambda'$ or $\Lambda''$, is compiling it into a classical planning task with conditional effects $P_{\Lambda}$, $P_{\Lambda'}$ or $P_{\Lambda''}$. The intuition behind these compilations is that a solution to the resulting classical planning task is a sequence of actions that:
\begin{enumerate}
\item Programs the STRIPS action model $\Xi$. For each $\xi\in\Xi$, the solution plan determines the literals that belong to the $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item Validates the programmed the STRIPS action model $\Xi$ using the given input knowledge (the given set of labels $\Sigma$, and $\Pi$ if available).  For every label {\small $1\leq t\leq \tau$}, then $\Xi$ is used to produce a final state $s_{n}^t$ starting from the corresponding initial state $s_0^t$. We call this process the validation of the programmed STRIPS $\Xi$ at the learning example {\small $1\leq t\leq \tau$}. If information about the corresponding plan $\pi_t$ is available, it is used to constrain the validation of $\Xi$.
\end{enumerate}

Figure~\ref{fig:plan} shows a classical plan for solving a learning task of the $\Lambda$ class. This classical plan programs and validates the operator schema {\tt\small stack}, from the blocksworld, using the plan $\pi_1$ and label $\sigma_1$ shown in Figure~\ref{fig:lexample}. In particular the steps $[0,8]$ are the actions for programming the preconditions of the operator, steps $[9,13]$ are the actions for programming the operator effects and steps $[14,22]$ are the actions for validating the programmed operator. The details of the compilation are given in the remaining of the section.

\begin{figure}[hbt]
\begin{scriptsize}
\begin{verbatim}
     0 : (program_pre_clear_stack_v1)
     1 : (program_pre_handempty_stack)
     2 : (program_pre_holding_stack_v2)
     3 : (program_pre_on_stack_v1_v1)
     4 : (program_pre_on_stack_v1_v2)
     5 : (program_pre_on_stack_v2_v1)
     6 : (program_pre_on_stack_v2_v2)
     7 : (program_pre_ontable_stack_v1)
     8 : (program_pre_ontable_stack_v2)
     9 : (program_eff_clear_stack_v1)
    10 : (program_eff_clear_stack_v2)
    11 : (program_eff_handempty_stack)
    12 : (program_eff_holding_stack_v1)
    13 : (program_eff_on_stack_v1_v2)
    14 : (apply_unstack a b i1 i2)
    15 : (apply_putdown a i2 i3)
    16 : (apply_unstack b c i3 i4)
    17 : (apply_stack b a i4 i5)
    18 : (apply_unstack c d i5 i6)
    19 : (apply_stack c b i6 i7)
    20 : (apply_pickup d i7 i8)
    21 : (apply_stack d c i8 i9)
    22 : (validate_1)
\end{verbatim}
\end{scriptsize}
 \caption{\small Example of a plan for programing and validating the operator schema $unstack$ using the plan $\pi_1$ and label $\sigma_1$ shown in Figure~\ref{fig:lexample} as well as previously specified operator schema for $pickup$, $putdown$ and $unstack$.}
\label{fig:plan}
\end{figure}

@@

Given a learning task $\Lambda''=\tup{\Psi,\Sigma}$ the compilation outputs a classical planning task $P_{\Lambda''}=\tup{F_{\Lambda''},A_{\Lambda''},I_{\Lambda''},G_{\Lambda''}}$ such that:
\begin{itemize}
\item $F_{\Lambda''}$ extends $F$ with:
\begin{itemize}
\item Fluents representing the programmed action model $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v$ and $\xi \in \Xi$). If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative effect/positive effect in the STRIPS operator schema $\xi\in \Xi$. 
\item A fluent $mode_{prog}$ indicating whether the operator schema are being programmed or they are already programmed operators that are being validated in the learning examples. 
\item Fluents $\{test_t\}_{1\leq t\leq \tau}$, indicating the learning example where the programmed action model is being validated.
\end{itemize}
\item $I_{\Lambda''}$, contains the fluents from $F$ that encode $s_0^1$ (i.e. the initial state of the first learning example), every $pre_f(\xi)\in F_{\Lambda}$ (note that our compilation assumes that initially, any operator schema, contains all the possible preconditions, no negative effect and no positive effect) and the $mode_{prog}$ set to true. 
\item $G_{\Lambda''}=\{test_t\}$, {\small $1\leq t\leq \tau$}, indicates that the programmed action model is validated in all the learning examples.
\item $A_{\Lambda''}$ contains actions of three different kinds:
\begin{enumerate}
\item The actions for programming the operator schema which includes:
\begin{itemize}
\item The actions for removing a {\em precondition} $f\in F_v$ from the action schema $\xi\in\Xi$.

@@

The binding of the operator schema is done implicitly by order of appearance of the action parameters, i.e. the variables in $pars(\xi)$ are bound to the objects in $\omega$ appearing in the same position. To illustrate this, Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying an already programmed operator schema $stack$.

@@

\begin{proof}[Proof sketch]
Once the preconditions of an operator schema $\Xi$ are programmed, they cannot be altered. The same happens with the positive and negative effects that define an operator schema $\xi \in \Xi$ (besides they can only be programmed after operator preconditions are programmed, because of the preconditions in the $programPre_{f,\xi}$ actions). Furthermore, once the operator schema are programmed they can only be applied because of the $mode_{prog}$ fluent. To solve the classical planning task $P_{\Lambda''}$, there is only one way of achieving a goal $\{test_t\}$,{\small $1\leq t\leq \tau$} that is by executing an applicable sequence of programmed operator schema that achieves the final state $s_n^t$ defined by the associated label $\sigma_t$, starting from the initial $s_0^t$ defined in this same label. If this is achieved for all the input examples of the learning task $\Lambda''$ (i.e. for all the labels $\sigma_t$, {\small $1\leq t\leq \tau$}), it means that the programmed action model $\Xi$ is compliant with all the input knowledge and hence, it is a solution to $\Lambda''$.
\end{proof}
